\pkg{sdmTMB}: An \proglang{R} Package for Fast, Flexible, and User-Friendly Generalized Linear Mixed Effects Models with Spatial and Spatiotemporal Random Fields

### 2023-10-19 doc/paper-jss/sdmTMB-paper.Rnw
fct fix

\begin{figure}[htbp]
\centering
\includegraphics[width=\textwidth]{../../figs/diagram}
\caption[]{
Components of an \sdmTMB\ model with illustrations, descriptions, examples, notation, and example code.
An \sdmTMB\ model can be built from any combination of the process components (first six rows) plus an observation component (last row).
The examples are from an SDM context, but the model can be fit to any spatially referenced point data.
Notation: We refer to design matrices as $\bm{X}$.
The indexes $\bm{s}$, $t$, and $g$ index spatial coordinates, time, and group, respectively.
The $\sigma$ and $\bm{\Sigma}$ symbols represent standard deviations and covariance matrices, respectively.
All other symbols refer to the  described model components (e.g., $\bm{\beta}$ and $\bm{\omega}$ refer to a vector of main effects and spatial random field deviations, respectively).
See the \sdmTMB\ \href{https://pbs-assess.github.io/sdmTMB/articles/}{model description vignette} for a full description of the model.
Note that \code{s()} denotes a smoother as in \pkg{mgcv} \citep{wood2017a}, \code{breakpt()} denotes a breakpoint `hockey-stick' shape \citep[e.g.,][]{barrowman2000}, \code{(1|g)} denotes a random intercept by group \code{g}, and \code{\textasciitilde 0} is used in an \R\ formula to omit an intercept.}\label{fig:diagram}
\end{figure}

@@

The expectation for a new data point is then the probability of a non-zero multiplied by the positive rate: $p \cdot r$.

@@

The \sdmTMB\ package is designed to be both modular and familiar to users of widely used \R\ packages (e.g., \pkg{glmmTMB}, \pkg{lme4}, \pkg{mgcv}).
The user starts by constructing a triangulation mesh for the SPDE approach with \fct{make\_mesh()} (Figure~\ref{fig:flowchart}).
\fct{make\_mesh()} is a wrapper for \pkg{fmesher} \citep{fmesher} triangulation mesh functions and users can also construct any mesh with \pkg{fmesher} or \INLA\ and pass it to \fct{make\_mesh()}.

Fitting is accomplished with \fct{sdmTMB()}, which has arguments similar to \pkg{glmmTMB}'s \fct{glmmTMB()} but with additional arguments for how any spatial and spatiotemporal random fields are structured, what column defines time, any time-varying formulas, and any spatially varying formulas.
Observation distributions and links are specified with standard \R\ family functions (e.g., \fct{binomial()}) plus several \sdmTMB-specific families not available in the \R\ stats library (e.g., \fct{nbinom2()}, \fct{delta\_lognormal()}, see \fct{?sdmTMB::Families}).

The \sdmTMB\ formula syntax (\code{formula} argument) follows a standard format that is similar to \pkg{glmmTMB}, \pkg{lme4}, and \pkg{mgcv}.
In addition to standard main effects, the user can include random intercepts (e.g., \code{+ (1 | group)}), threshold shaped hockey-stick models \fct{breakpt()} \citep{barrowman2000}, logistic functions \fct{logistic()}, and penalized smoothers \fct{s()} for generalized additive models, GAMS \citep{wood2017a}.
Penalized smoothers use the same \fct{s()} and \fct{t2()} syntax as in \pkg{mgcv} \citep{wood2017a}.
Supported functionality includes bivarate smoothers \fct{s(x, y)},
smoothers varying by continuous or categorical variables \fct{s(x1, by = x2)},
cyclical smoothers \fct{s(x, bs = "cc")},
and smoothers with specified basis dimensions \fct{s(x, k = 4)} \citep{wood2017a}.
Beyond the main formula, \fct{sdmTMB()} accepts one-sided formulas for coefficients that should vary through time (\code{time\_varying}) according to a random walk or AR(1) process (\code{time\_varying\_type}) or vary through space as random fields (\code{spatial\_varying}).

Once the user makes a call to \fct{sdmTMB()}, input data structures for a \TMB\ model are constructed internally (Figure~\ref{fig:flowchart}).
If needed, data structures required to implement penalized smoothers are constructed with \fct{smooth2random()} from \pkg{mgcv} \citep{wood2017a}, similarly to \pkg{gamm4} \citep{wood2020} and \pkg{brms} \citep{brms}.
\fct{sdmTMB()} formats data, establishes parameter starting values, and constructs an objective function with derivatives based on a compiled \proglang{C++} template written for \TMB.
The objective function returns the marginal log likelihood and its gradient, integrating over random effects with the Laplace approximation \citep{kristensen2016}.
The negative marginal log likelihood is minimized via the non-linear optimization routine \fct{stats::nlminb()} \citep{gay1990} in \R\ with additional optimization carried out via a Newton optimization step to find the Hessian using \fct{stats::optimHess()} \citep{gay1990, r2021}.
After rapid model exploration with maximum likelihood, one can optionally pass an \sdmTMB\ model to the \R\ package \pkg{rstan} \citep{carpenter2017, rstan} via \pkg{tmbstan} \citep{monnahan2018} to samples from the joint posterior distribution for Bayesian inference.

A fitted model summary can be viewed with \fct{print()} or \fct{summary()} and a set of basic ``sanity'' checks can be run with \fct{sanity()}.
\fct{tidy()} returns parameter estimates in standard data frame formats similar to \pkg{broom} \citep{broom}.
Others standard S3 methods are also available such as \fct{fixef()}, \fct{confint()}, and \fct{vcov()}.

Prediction on fitted or new data is accomplished with \fct{predict()} (\fct{?predict.sdmTMB}).
The \fct{predict.sdmTMB()} method is flexible and includes the option to specify a new data frame (\fct{newdata}),
whether to return predictions on the link or response scale (type),
whether to return standard errors (\code{se\_fit},
can be slow if conditioned on random fields),
whether to condition on the random fields (\code{re\_form}),
whether to condition on the random intercepts (\code{re\_form\_iid}),
which delta model linear predictor to use (\code{model}),
whether to take draws from the joint parameter precision matrix (\code{nsim}),
and whether to use MCMC samples from a \pkg{tmbstan} model fit (\code{mcmc\_samples}).

A variety of model evaluation tools are available.
A \fct{residuals()} method calculates various types of residuals.
New observations can be simulated from a fitted model with \fct{simulate()} or new data can be simulated without a fitted model (``de novo'') with \fct{sdmTMB\_simulate()}.
\fct{sdmTMB\_cv()} facilitates cross validation.

% Derived quantities can be calculated on the fitted model.
% TODO: more COG refs
% Common quantities of interest include summarizing shifts in the center of gravity (\fct{get\_cog()}) \citep[e.g.,][]{thorson2016cog}, or tracking total densities through time (\fct{get\_index()}).

% The \fct{get\_index()} function generates a weighted average of densities for the prediction grid, integrating across space; for prediction grids with equally sized cells, each cell will contribute equally to the total average, however cells may be of unequal size.

% Data may be simulated with or without a fitted model (\fct{simulate.sdmTMB()}, \fct{sdmTMB\_simulate()} respectively).
% Estimation is done via maximum marginal likelihood with the \fct{sdmTMB()} function, and the resulting fitted model can be used to make predictions (\fct{predict.sdmTMB()}) or calculate residuals (\fct{residuals.sdmTMB()}), derive additional quantities of interest, and simulate Bayesian posterior distributions.
% Common quantities of interest may include summarizing shifts in the center of gravity (\fct{get\_cog()}) \citep[e.g.,][]{thorson2016cog}, or tracking total densities through time (\fct{get\_index()}).
% The \fct{get\_index()} function generates a weighted average of densities for the prediction grid, integrating across space; for prediction grids with equally sized cells, each cell will contribute equally to the total average, however cells may be of unequal size.

\sdmTMB\ models can include penalized likelihoods by assigning priors (penalties) to model parameters through the \fct{sdmTMB()} \code{prior} argument (\code{?sdmTMBpriors}).
These priors may be useful in cases where estimation is difficult because of identifiability issues or relatively flat likelihood surfaces, or to impart prior information or achieve regularization.
Following other recent SPDE implementations in \TMB\ \citep{breivik2021, osgood-zimmerman2021}, penalized complexity (PC) priors \citep{simpson2017, fuglstad2019} (\code{?pc\_matern}) can constrain the spatial range and variance parameters.
These priors or penalties are available both with maximum likelihood estimation and with MCMC sampling.
If a model will be passed to \pkg{tmbstan} with priors, a \code{bayesian} logical argument should be set to \code{TRUE} to enable Jacobian adjustments for changes of variables (priors applied to parameters that are internally transformed) \citep{carpenter2017}.

@@

% Validation and selection of state-space models is challenging, particularly when
% using the Laplace approximation \citep{thygesen2017a}. We provide several
% approaches to assist this process: (1) The Akaike Information Criterion
% \citep[AIC,][]{akaike1974} can be calculated with \fct{AIC()}, although AIC
% has well-documented biases with mixed-effects models \citep{liang2008}. (2)
% Alternatively, k-fold cross validation with \fct{sdmTMB\_cv()} can be used
% with user-specified or randomly chosen folds for model selection or to evaluate
% goodness of fit according to user-calculated criteria (e.g., mean squared error,
% area under the curve). (3) An \sdmTMB\ model can be passed to the
% \pkg{tmbstan} package \citep{monnahan2018} to sample from the joint posterior
% with Stan \citep{carpenter2017}, evaluate the accuracy of the Laplace
% approximation, or perform posterior predictive checks (see
% \code{?extract\_mcmc}). (4) The \fct{residuals()} method by default returns
% randomized quantile \citep{dunn1996} or probability integral transform (PIT)
% \citep{smith1985} residuals. For state-space models, these residuals have known
% statistical issues with the Laplace approximation \citep{thygesen2017a} but are
% quick to calculate. A version that uses Markov chain Monte Carlo (MCMC) sampling
% of the random effects to avoid this issue \citep{rufener2021} is recommended but
% slower (\code{?residuals.sdmTMB}). Simulation-based residuals \citep{dharma}
% are also possible. (5) The \fct{simulate.sdmTMB()} method can simulate from
% fitted models and the \fct{sdmTMB\_simulate()} function facilitates
% simulating data without starting from a fitted model. Models can be fit to these
% simulated data to ensure identifiability, evaluate bias and precision in
% parameter estimation, or evaluate the consequences of model misspecification.

@@

% \sdmTMB\ relies on several well-established \R\ packages (Figure~\ref{fig:flowchart}).
% \sdmTMB\ uses \pkg{fmesher} \citep{fmesher} and \INLA\ \citep{lindgren2015} to generate the triangulation mesh and input matrices for the SPDE calculations.
% The SPDE precision matrix calculations themselves are performed by functions within \TMB\ \citep{kristensen2016} (\code{R\_inla} namespace).
% Data structures required to implement penalized smoothers are constructed with \fct{smooth2random()} from \pkg{mgcv} \citep{wood2017a}, similarly to \pkg{gamm4} \citep{wood2020} and \pkg{brms} \citep{brms}.
% The statistical model itself is written as a \proglang{C++} template for \TMB.
% \TMB\ returns the marginal log likelihood and its gradient, integrating over random effects with the Laplace approximation \citep{kristensen2016}.
% The negative marginal log likelihood is minimized via the non-linear optimization routine \fct{stats::nlminb()} \citep{gay1990} in \R\ with additional optimization carried out via a Newton optimization step to find the Hessian using \fct{stats::optimHess()} \citep{gay1990, r2021}.
% After rapid model exploration with maximum likelihood, one can optionally pass an \sdmTMB\ model to the \R\ package \pkg{rstan} \citep{carpenter2017, rstan} via \pkg{tmbstan} \citep{monnahan2018} to samples from the joint posterior distribution for Bayesian inference.
% % TODO: visreg + ggeffects effects, etc.?

@@

\sdmTMB\ can be installed from the Comprehensive \R
Archive Network (CRAN) at
\url{https://CRAN.R-project.org/package=sdmTMB}

@@

Additional utilities are maintained in the \pkg{sdmTMBextra} package at\\
\url{https://github.com/pbs-assess/sdmTMBextra}

@@

An \sdmTMB\ model requires a data frame that contains a response column, columns for any predictors, and columns for spatial coordinates.
Usually it makes sense to convert the spatial coordinates to an equidistant projection such as UTMs to ensure that distance remains constant throughout the study region \citep[e.g., using \fct{sf::st\_transform()},][]{pebesma2018}.
Here we use the helper function \fct{add\_utm\_columns()} to add UTM coordinates with km units (so our estimated spatial range parameter is not too big or small).
By default, the function will guess the UTM zone and create new columns \code{X} and \code{Y}.
Since our example data already has these UTM columns, we can skip running this code.

@@

\sloppy We then create a mesh object that contains triangulation and projection matrices needed to apply the SPDE approach using \fct{make\_mesh()}.
% \fct{make\_mesh()} is a wrapper function for functions in the \pkg{fmesher} package.
The argument \code{cutoff} defines the minimum allowed distance between mesh vertices in the units of \code{X} and \code{Y} (km).
We could create a basic mesh just specifying this:

@@

We can also specify additional arguments, in this case passed to \fct{fmesher::fm\_mesh\_2d\_inla()}: a maximum triangle edge length of 10 km and 40 km for the inner and outer mesh, and an offset width of 10 km and 40 km for the inner and outer meshes borders.
% Alternatively, we could have created any mesh via \INLA\ or \pkg{fmesher} and supplied it to the \code{mesh} argument in \fct{make\_mesh()}.
For more irregularly shaped areas, we could have specified a non-convex hull with \code{convex} and \code{concave} arguments.
The triangle edge length should be several times smaller than the range and the outer boundary should extend at least as far as the range (REF).

@@

<<dog-binomial-mesh2, results='hide', message=FALSE, warning=FALSE, echo=TRUE, fig.cap="SPDE mesh (lines) combined with the trawl survey observations (points). The locations where lines intersect are referred to as ``vertices''. Finer meshes will be slower to fit but generally increase the accuracy of the SPDE approximation. A greater degree of control over the mesh construction can be achieved by using \\pkg{fmesher} or \\proglang{R}-\\pkg{INLA} directly and supplying the resulting object to \\code{make\\_mesh()}.", fig.width=4.5, fig.asp=1, out.width="3in">>=
mesh_pcod2 <- make_mesh(
  pcod, c("X", "Y"),
  fmesher_func = fmesher::fm_mesh_2d_inla,
  cutoff = 8, # minimum triangle edge length
  max.edge = c(10, 40), # max triangle edge length; inner and outer meshes
  offset = c(10, 40) # inner and outer border widths
)
plot(mesh_pcod2)
@

@@

We can implement this model with \fct{sdmTMB()}:

@@

We can also fit a version that omits the spatial random field by setting \code{spatial = "off"}.
We will use the \fct{update()} S3 method to refit the model while updating any specified arguments:

@@

We can run some basic checks on our model with the \fct{sanity()} function:

@@

This does not flag any issues. \fct{sanity()} is checking that the \fct{nlminb()} optimizer reported successful convergence,
that the Hessian matrix is positive definite,
that no extreme or small eigenvalues are detected,
that no absolute log likelihood gradients with respect to fixed defects are $\le$ 0.001,
that all fixed effects have reported standard errors that do not look unreasonably large ($<$ 100 by default),
that random field marginal standard deviations are not unexpectedly small or large ($>$ 0.01 and $<$ 100),
and that the random field Matérn range parameter does not look unreasonably large ($<$ 1.5 times the largest edge of a bounding box around the observations).

@@

We can use the \fct{tidy()} function to obtain a data frame with parameter estimates (standard methods such as \fct{fixef()}, \fct{confint()}, and \fct{vcov()} are also available).
The standard errors on our fixed effects have increased with the spatial random field:

@@

By setting \code{effects = "ran\_pars"}, \fct{tidy()} will return random
field parameters, where \code{sigma\_O} is the marginal standard deviation of the spatial random field \(\bm{\omega}\) (`O' for `Omega').

@@

We can test for spatial autocorrelation with a Moran's I test or with a visual inspection of the residuals calculated from the \fct{residuals()} S3 method.
We omit example code for Moran's I calculation for brevity but include it in the code supplement.
% TODO: reword this?
We see strong evidence for spatial autocorrelation for the model without a random field (p $<$ 0.01) but no evidence for the model with a random field (p = 0.84) suggesting that residual spatial autocorrelation is alleviated by including the random field.

@@

We can do additional model comparison with cross validation via \fct{sdmTMB\_cv()}.
We will do 10-fold cross validation with the folds constructed randomly.
Using the \code{fold\_ids} argument, we could supply our own folds and conduct spatially blocked cross validation. If we set a parallel plan with the \pkg{future} package \citep{bengtsson2021future}, our folds will be fit in parallel.

@@

\sloppy To visualize our model, we can make predictions with the \fct{predict()} method (\code{?predict.sdmTMB}) and optionally use the \code{newdata} argument to predict on a new data frame containing locations and all predictor columns.
Here, we will predict on a 2 \(\times\) 2 km grid (\code{qcs\_grid}) that covers the entire region of interest so we can visualize the predictions spatially.
The grid contains spatial covariate columns and all predictors used in the model set at values for which we want to predict.
The output of \fct{predict()} is a data frame containing overall estimates in link space (\code{est}), estimates from the non-random-field components (\code{est\_non\_rf}; here, intercept and depth), and estimates from the individual random field components (here, \code{omega\_s}---the spatial field).
We can plot these with \fct{geom\_raster()} from the \pkg{ggplot2} package (Figure~\ref{fig:pcod-predict-maps}).

@@

We find that the anisotropic AR(1) is favoured. It makes sense that anisotropy is important here given the elongated shape of the continental shelf with a rapid transition to deeper water. We can use \fct{plot\_anisotropy()} to visually inspect the anisotropy (Figure~\ref{fig:dog-aniso}).

@@

On inspecting the model output with \fct{summary()}, we notice that the spatial random field SD in the positive model component has become very small once AR(1) spatiotemporal fields were added.
We therefore consider a model where these are turned off in the second component of the delta model and choose this model.

@@

We save our chosen model to the object \code{fit} to simplify subsequent code, run the \fct{sanity()} check (suppressed for brevity), and inspect \fct{summary()}:

@@

Similarly to the first example, we can visualize model predictions on a grid covering the area of interest.
Because this is a spatiotemporal model, we first need to replicate our grid for each year we will predict on.
Since this is a common operation, we include the function \fct{replicate\_df()} to replicate a data frame.
We then ensure our data frame contains all the predictors used in the model (here \code{log\_depth}).

@@

\fct{summary()} prints standard model information:

>>>>

\begin{figure}[htbp]
\centering
\includegraphics[width=\textwidth]{../../figs/diagram}
\caption[]{
Components of an \sdmTMB\ model with illustrations, descriptions, examples, notation, and example code.
An \sdmTMB\ model can be built from any combination of the process components (first six rows) plus an observation component (last row).
The examples are from an ecology context, but the model can be fit to any spatially referenced point data.
Notation: We refer to design matrices as $\bm{X}$.
The indexes $\bm{s}$, $t$, and $g$ index spatial coordinates, time, and group, respectively.
The $\sigma$ and $\bm{\Sigma}$ symbols represent standard deviations and covariance matrices, respectively.
All other symbols refer to the  described model components (e.g., $\bm{\beta}$ and $\bm{\omega}$ refer to a vector of main effects and spatial random field deviations, respectively).
Note that \code{s()} denotes a smoother as in \pkg{mgcv} \citep{wood2017a}, \code{breakpt()} denotes a breakpoint ``hockey-stick'' shape \citep[e.g.,][]{barrowman2000}, \code{(1|g)} denotes a random intercept by group \code{g}, and \code{\textasciitilde 0} is used in an \R\ formula to omit the intercept.}\label{fig:diagram}
\end{figure}

@@

The expectation for a new data point is then the probability of a non-zero event multiplied by the positive rate: $p \cdot r$.

@@

The \sdmTMB\ package is designed to be both modular and familiar to users of widely used \R\ packages (e.g., \pkg{glmmTMB}, \pkg{lme4}, \pkg{mgcv}).
The user starts by constructing a triangulation mesh for the SPDE approach with \fct{make\_mesh} (Figure~\ref{fig:flowchart}).
\fct{make\_mesh} is a wrapper for \pkg{fmesher} \citep{fmesher} triangulation mesh functions and users can also construct any mesh with \pkg{fmesher} or \INLA\ and pass it to \fct{make\_mesh}.

Fitting is accomplished with \fct{sdmTMB}, which has arguments similar to \pkg{glmmTMB}'s \fct{glmmTMB} but with additional arguments for how any spatial and spatiotemporal random fields are structured, what column defines time, any time-varying formulas, and any spatially varying formulas.
Observation distributions and links are specified with standard \R\ family functions (e.g., \fct{binomial}) plus several \sdmTMB-specific families not available in the \R\ stats library (e.g., \fct{nbinom2}, \fct{delta\_lognormal}, see \code{?sdmTMB::Families}).

The \sdmTMB\ formula syntax (\code{formula} argument) follows a standard format that is similar to \pkg{glmmTMB}, \pkg{lme4}, and \pkg{mgcv}.
In addition to standard main effects, the user can include random intercepts (e.g., \code{+ (1 | group)}), threshold shaped hockey-stick models \fct{breakpt} \citep{barrowman2000}, logistic functions \fct{logistic}, and penalized smoothers \fct{s} for generalized additive models, GAMS \citep{wood2017a}.
Penalized smoothers use the same \fct{s} and \fct{t2} syntax as in \pkg{mgcv} \citep{wood2017a}.
Supported functionality includes bivarate smoothers \code{s(x, y)},
smoothers varying by continuous or categorical variables \code{s(x1, by = x2)},
cyclical smoothers \code{s(x, bs = "cc")},
and smoothers with specified basis dimensions \code{s(x, k = 4)} \citep{wood2017a}.
Beyond the main formula, \fct{sdmTMB} accepts one-sided formulas for coefficients that should vary through time (\code{time\_varying}) according to a random walk or AR(1) process (\code{time\_varying\_type}) or vary through space as random fields (\code{spatial\_varying}).

Once the user makes a call to \fct{sdmTMB}, input data structures for a \TMB\ model are constructed internally (Figure~\ref{fig:flowchart}).
If needed, data structures required to implement penalized smoothers are constructed with \fct{smooth2random} from \pkg{mgcv} \citep{wood2017a}, similarly to \pkg{gamm4} \citep{wood2020} and \pkg{brms} \citep{brms}.
\fct{sdmTMB} formats data, establishes parameter starting values, and constructs an objective function with derivatives based on a compiled \proglang{C++} template written for \TMB.
The objective function returns the marginal log likelihood and its gradient, integrating over random effects with the Laplace approximation \citep{kristensen2016}.
The negative marginal log likelihood is minimized via the non-linear optimization routine \fct{stats::nlminb} \citep{gay1990} in \R\ with additional optimization carried out via a Newton optimization step to find the Hessian using \fct{stats::optimHess} \citep{gay1990, r2021}.
After rapid model exploration with maximum likelihood, one can optionally pass an \sdmTMB\ model to the \R\ package \pkg{rstan} \citep{carpenter2017, rstan} via \pkg{tmbstan} \citep{monnahan2018} to samples from the joint posterior distribution for Bayesian inference.

A fitted model summary can be viewed with \fct{print} or \fct{summary} and a set of basic ``sanity'' checks can be run with \fct{sanity}.
\fct{tidy} returns parameter estimates in standard data frame formats similar to \pkg{broom} \citep{broom}.
Others standard S3 methods are also available such as \fct{fixef}, \fct{confint}, and \fct{vcov}.

Prediction on fitted or new data is accomplished with \fct{predict} (\code{?predict.sdmTMB}).
The predict method is flexible and includes the option to specify a new data frame (\code{newdata}),
whether to return predictions on the link or response scale (type),
whether to return standard errors (\code{se\_fit},
can be slow if conditioned on random fields),
whether to condition on the random fields (\code{re\_form}),
whether to condition on the random intercepts (\code{re\_form\_iid}),
which delta model linear predictor to use (\code{model}),
whether to take draws from the joint parameter precision matrix (\code{nsim}),
and whether to use MCMC samples from a \pkg{tmbstan} model fit (\code{mcmc\_samples}).

A variety of model evaluation tools are available.
A \fct{residuals} method calculates various types of residuals.
New observations can be simulated from a fitted model with \fct{simulate} or new data can be simulated without a fitted model (``de novo'') with \fct{sdmTMB\_simulate}.
\fct{sdmTMB\_cv} facilitates cross validation.

% Derived quantities can be calculated on the fitted model.
% TODO: more COG refs
% Common quantities of interest include summarizing shifts in the center of gravity (\fct{get\_cog}) \citep[e.g.,][]{thorson2016cog}, or tracking total densities through time (\fct{get\_index}).

% The \fct{get\_index} function generates a weighted average of densities for the prediction grid, integrating across space; for prediction grids with equally sized cells, each cell will contribute equally to the total average, however cells may be of unequal size.

% Data may be simulated with or without a fitted model (\fct{simulate.sdmTMB}, \fct{sdmTMB\_simulate} respectively).
% Estimation is done via maximum marginal likelihood with the \fct{sdmTMB} function, and the resulting fitted model can be used to make predictions (\fct{predict.sdmTMB}) or calculate residuals (\fct{residuals.sdmTMB}), derive additional quantities of interest, and simulate Bayesian posterior distributions.
% Common quantities of interest may include summarizing shifts in the center of gravity (\fct{get\_cog}) \citep[e.g.,][]{thorson2016cog}, or tracking total densities through time (\fct{get\_index}).
% The \fct{get\_index} function generates a weighted average of densities for the prediction grid, integrating across space; for prediction grids with equally sized cells, each cell will contribute equally to the total average, however cells may be of unequal size.

\sdmTMB\ models can include penalized likelihoods by assigning priors (penalties) to model parameters through the \fct{sdmTMB} \code{prior} argument (\code{?sdmTMBpriors}).
These priors may be useful in cases where estimation is difficult because of identifiability issues or relatively flat likelihood surfaces, or to impart prior information or achieve regularization.
Following other recent SPDE implementations in \TMB\ \citep{breivik2021, osgood-zimmerman2021}, penalized complexity (PC) priors \citep{simpson2017, fuglstad2019} (\code{?pc\_matern}) can constrain the spatial range and variance parameters.
These priors or penalties are available both with maximum likelihood estimation and with MCMC sampling.
If a model will be passed to \pkg{tmbstan} with priors, a \code{bayesian} logical argument should be set to \code{TRUE} to enable Jacobian adjustments for changes of variables (priors applied to parameters that are internally transformed) \citep{carpenter2017}.

@@

% Validation and selection of state-space models is challenging, particularly when
% using the Laplace approximation \citep{thygesen2017a}. We provide several
% approaches to assist this process: (1) The Akaike Information Criterion
% \citep[AIC,][]{akaike1974} can be calculated with \fct{AIC}, although AIC
% has well-documented biases with mixed-effects models \citep{liang2008}. (2)
% Alternatively, k-fold cross validation with \fct{sdmTMB\_cv} can be used
% with user-specified or randomly chosen folds for model selection or to evaluate
% goodness of fit according to user-calculated criteria (e.g., mean squared error,
% area under the curve). (3) An \sdmTMB\ model can be passed to the
% \pkg{tmbstan} package \citep{monnahan2018} to sample from the joint posterior
% with Stan \citep{carpenter2017}, evaluate the accuracy of the Laplace
% approximation, or perform posterior predictive checks (see
% \code{?extract\_mcmc}). (4) The \fct{residuals} method by default returns
% randomized quantile \citep{dunn1996} or probability integral transform (PIT)
% \citep{smith1985} residuals. For state-space models, these residuals have known
% statistical issues with the Laplace approximation \citep{thygesen2017a} but are
% quick to calculate. A version that uses Markov chain Monte Carlo (MCMC) sampling
% of the random effects to avoid this issue \citep{rufener2021} is recommended but
% slower (\code{?residuals.sdmTMB}). Simulation-based residuals \citep{dharma}
% are also possible. (5) The \fct{simulate.sdmTMB} method can simulate from
% fitted models and the \fct{sdmTMB\_simulate} function facilitates
% simulating data without starting from a fitted model. Models can be fit to these
% simulated data to ensure identifiability, evaluate bias and precision in
% parameter estimation, or evaluate the consequences of model misspecification.

@@

% \sdmTMB\ relies on several well-established \R\ packages (Figure~\ref{fig:flowchart}).
% \sdmTMB\ uses \pkg{fmesher} \citep{fmesher} and \INLA\ \citep{lindgren2015} to generate the triangulation mesh and input matrices for the SPDE calculations.
% The SPDE precision matrix calculations themselves are performed by functions within \TMB\ \citep{kristensen2016} (\code{R\_inla} namespace).
% Data structures required to implement penalized smoothers are constructed with \fct{smooth2random} from \pkg{mgcv} \citep{wood2017a}, similarly to \pkg{gamm4} \citep{wood2020} and \pkg{brms} \citep{brms}.
% The statistical model itself is written as a \proglang{C++} template for \TMB.
% \TMB\ returns the marginal log likelihood and its gradient, integrating over random effects with the Laplace approximation \citep{kristensen2016}.
% The negative marginal log likelihood is minimized via the non-linear optimization routine \fct{stats::nlminb} \citep{gay1990} in \R\ with additional optimization carried out via a Newton optimization step to find the Hessian using \fct{stats::optimHess} \citep{gay1990, r2021}.
% After rapid model exploration with maximum likelihood, one can optionally pass an \sdmTMB\ model to the \R\ package \pkg{rstan} \citep{carpenter2017, rstan} via \pkg{tmbstan} \citep{monnahan2018} to samples from the joint posterior distribution for Bayesian inference.
% % TODO: visreg + ggeffects effects, etc.?

@@

\sdmTMB\ can be installed from the Comprehensive \R\ Archive Network (CRAN) at
\url{https://CRAN.R-project.org/package=sdmTMB}

@@

Additional utilities that require heavier package dependencies (such as \INLA\ and \pkg{rstan}) are maintained in the \pkg{sdmTMBextra} package at\\
\url{https://github.com/pbs-assess/sdmTMBextra}

@@

An \sdmTMB\ model requires a data frame that contains a response column, columns for any predictors, and columns for spatial coordinates.
Usually it makes sense to convert the spatial coordinates to an equidistant projection such as UTMs to ensure that distance remains constant throughout the study region \citep[e.g., using \fct{sf::st\_transform},][]{pebesma2018}.
Here we use the helper function \fct{add\_utm\_columns} to add UTM coordinates with km units (so our estimated spatial range parameter is not too big or small).
By default, the function will guess the UTM zone and create new columns \code{X} and \code{Y}.
Since our example data already has these UTM columns, we can skip running this code.

@@

\sloppy We then create a mesh object that contains triangulation and projection matrices needed to apply the SPDE approach using \fct{make\_mesh}.
% \fct{make\_mesh} is a wrapper function for functions in the \pkg{fmesher} package.
The argument \code{cutoff} defines the minimum allowed distance between mesh vertices in the units of \code{X} and \code{Y} (km).
We could create a basic mesh just specifying this:

@@

We can also specify additional arguments, in this case passed to \fct{fmesher::fm\_mesh\_2d\_inla}: a maximum triangle edge length of 10 km and 40 km for the inner and outer mesh, and an offset width of 10 km and 40 km for the inner and outer meshes borders.
% Alternatively, we could have created any mesh via \INLA\ or \pkg{fmesher} and supplied it to the \code{mesh} argument in \fct{make\_mesh}.
For more irregularly shaped areas, we could have specified a non-convex hull with \code{convex} and \code{concave} arguments.
The triangle edge length should be several times smaller than the range and the outer boundary should extend at least as far as the range (REF).

@@

<<dog-binomial-mesh2, results='hide', message=FALSE, warning=FALSE, echo=TRUE, fig.cap="SPDE mesh (lines) combined with the trawl survey observations (points). The locations where lines intersect are referred to as ``vertices''. Finer meshes will be slower to fit but generally increase the accuracy of the SPDE approximation. A greater degree of control over the mesh construction can be achieved by using \\pkg{fmesher} or \\proglang{R}-\\pkg{INLA} directly and supplying the object to \\code{make\\_mesh()}.", fig.width=4.5, fig.asp=1, out.width="3in">>=
mesh_pcod2 <- make_mesh(
  pcod, c("X", "Y"),
  fmesher_func = fmesher::fm_mesh_2d_inla,
  cutoff = 8, # minimum triangle edge length
  max.edge = c(10, 40), # max triangle edge length; inner and outer meshes
  offset = c(10, 40) # inner and outer border widths
)
plot(mesh_pcod2)
@

@@

We can implement this model with \fct{sdmTMB}:

@@

We can also fit a version that omits the spatial random field by setting \code{spatial = "off"}.
We will use the \fct{update} S3 method to refit the model while updating any specified arguments:

@@

We can run some basic checks on our model with the \fct{sanity} function:

@@

This does not flag any issues. \fct{sanity} is checking that the \fct{nlminb} optimizer reported successful convergence,
that the Hessian matrix is positive definite,
that no extreme or small eigenvalues are detected,
that no absolute log likelihood gradients with respect to fixed defects are $\le$ 0.001,
that all fixed effects have reported standard errors that do not look unreasonably large ($<$ 100 by default),
that random field marginal standard deviations are not unexpectedly small or large ($>$ 0.01 and $<$ 100),
and that the random field Matérn range parameter does not look unreasonably large ($<$ 1.5 times the largest edge of a bounding box around the observations).

@@

We can use the \fct{tidy} function to obtain a data frame with parameter estimates (standard methods such as \fct{fixef}, \fct{confint}, and \fct{vcov} are also available).
The standard errors on our fixed effects have increased with the spatial random field:

@@

By setting \code{effects = "ran\_pars"}, \fct{tidy} will return random
field parameters, where \code{sigma\_O} is the marginal standard deviation of the spatial random field \(\bm{\omega}\) (`O' for `Omega').

@@

We can test for spatial autocorrelation with a Moran's I test or with a visual inspection of the residuals calculated from the \fct{residuals} S3 method.
We omit example code for Moran's I calculation for brevity but include it in the code supplement.
% TODO: reword this?
We see strong evidence for spatial autocorrelation for the model without a random field (p $<$ 0.01) but no evidence for the model with a random field (p = 0.84) suggesting that residual spatial autocorrelation is alleviated by including the random field.

@@

We can do additional model comparison with cross validation via \fct{sdmTMB\_cv}.
We will do 10-fold cross validation with the folds constructed randomly.
Using the \code{fold\_ids} argument, we could supply our own folds and conduct spatially blocked cross validation. If we set a parallel plan with the \pkg{future} package \citep{bengtsson2021future}, our folds will be fit in parallel.

@@

\sloppy To visualize our model, we can make predictions with the \fct{predict} method (\code{?predict.sdmTMB}) and optionally use the \code{newdata} argument to predict on a new data frame containing locations and all predictor columns.
Here, we will predict on a 2 \(\times\) 2 km grid (\code{qcs\_grid}) that covers the entire region of interest so we can visualize the predictions spatially.
The grid contains spatial covariate columns and all predictors used in the model set at values for which we want to predict.
The output of \fct{predict} is a data frame containing overall estimates in link space (\code{est}), estimates from the non-random-field components (\code{est\_non\_rf}; here, intercept and depth), and estimates from the individual random field components (here, \code{omega\_s}---the spatial field).
We can plot these with \fct{geom\_raster} from the \pkg{ggplot2} package (Figure~\ref{fig:pcod-predict-maps}).

@@

We find that the anisotropic AR(1) is favoured. It makes sense that anisotropy is important here given the elongated shape of the continental shelf with a rapid transition to deeper water. We can use \fct{plot\_anisotropy} to visually inspect the anisotropy (Figure~\ref{fig:dog-aniso}).

@@

On inspecting the model output with \fct{summary}, we notice that the spatial random field SD in the positive model component has become very small once AR(1) spatiotemporal fields were added.
We therefore consider a model where these are turned off in the second component of the delta model and choose this model.

@@

We save our chosen model to the object \code{fit} to simplify subsequent code, run the \fct{sanity} check (suppressed for brevity), and inspect \fct{summary}:

@@

Similarly to the first example, we can visualize model predictions on a grid covering the area of interest.
Because this is a spatiotemporal model, we first need to replicate our grid for each year we will predict on.
Since this is a common operation, we include the function \fct{replicate\_df} to replicate a data frame.
We then ensure our data frame contains all the predictors used in the model (here \code{log\_depth}).

@@

\fct{summary} prints standard model information:
